---
title: "Classification_tree_test"
author: "Jiashu Liu"
date: "2023-12-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(palmerpenguins)
library(tidyverse)
library(dplyr)
library(stringr)
data("penguins")
# Classification question: species classification
# Predict the species of the penguin (e.g., Adelie, Chinstrap, Gentoo) based on other variables such as bill length, bill depth, flipper length, body mass, island, and sex.
```

```{r}
# Preprocessing Data
penguins <- na.omit(penguins)
penguins <- penguins %>% 
  select(-year)
#penguins$island <- as.numeric(as.factor(penguins$island))
#penguins$sex <- as.numeric(as.factor(penguins$sex))
```

Thoughts on building a simple classification tree:

1. Define a function for gini_impurity
2. Loop over all the predictors and possible split points. Pick the predictor that result in the lowest weighted average Gini impurity.

OVERALL, there are 4 numerical variables and 2 categorical variables with at most 333 observations
2.1 for numeric predictors, we should calculate the average for all adjacent penguins, and then calcualte the Gini impurity for each average numeric value.
Choose the threshold that has the lowest weighted total Gini impurity
2.2 for cateorical predictors, just calculate the Gini impurity

3. Starting from the previously formed regions, repeat step 2. 
4. Stop once no possible split can lower a node's Gini impurity. 

```{r}
numeric_features <- names(penguins)[sapply(penguins, is.numeric)]
categorical_features <- names(penguins)[sapply(penguins, is.factor)] 

print("Numeric Features:")
print(numeric_features)

print("Categorical Features:")
print(categorical_features)
```
```{r}
# Calculate Gini Index (Classification)
gini_impurity <- function(y){
  # assumes y if a factor with all levels
  if(length(y) == 0) return(0)
  p <- table(y)/length(y)
  1-sum(p^2)
}
# Entropy (Classification)
entropy <- function(y){
  # assumes y is a factor with all levels
  if(length(y)==0) return(0)
  p <- table(y)/length(y)
  sum(-p*log2(p+1e-9))
}
# Regression
variance <- function(y){
  if(length(y) <= 1) return(0)
  var(y)
}
```

```{r}

```



















